Grade Retention and Mixed Race

### 2022-08-15 products/paper/main.qmd
Add first draft of introduction

---
title: Grade Retention and Mixed Race
author: Aaron Gullickson, *University of Oregon*
abstract: This is a test abstract
thanks: Thanks to people and stuff
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
    #template: resources/aog-latex-ms.tex
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

@@

@@

@@

## Conclusions

## References

>>>>

---
title: Grade Retention and Mixed Race
author: 
  - Aaron Gullickson, *University of Oregon*
abstract: This is a test abstract
thanks: The corresponding author can be reached at aarong@uoregon.edu. All code and data for this project are available at ...
format:
  pdf:
    documentclass: scrartcl
    papersize: letter
    fig-height: 4
    fig-pos: t
    keep-tex: false
  docx:
    toc: false
bibliography: ../project.bib
execute: 
  eval: true
  echo: false
  warning: false
  error: false
editor: source
---

@@

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==). This demographic phenomenon may have a potentially transformative effect on racial boundaries. Although race mixing has always been a feature of US history (and the history of any racialized social system) (==CITE==), this biracial boom is notable for its demographic size and for occurring in  post-Civil Rights ear in which historical norms govern how to handle racial mixture may be dissolving or shifting. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into the prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==). ==SUMMARIZE BRIEFLY==. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers. The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.

First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic. Research on the identification and classification of biracial individuals indicates that many individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==). Social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials. Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-classification.

This problem of reverse causation can be largely addressed when researchers have access to the racial identification of a respondent's biological parents. While such information is available on some large scale longitudinal surveys, researchers then run into a second problem of sample size. The samples of biracial respondents in such cases are generally far too small to make reliable statistical inferences. This limitation is particularly severe for biracial individuals with two non-white parents.

In this article, I exploit features of the American Community Survey to identify one life outcome that is available in large scale Census data - grade retention. Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==). Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience in their life and within it we can clearly see the imprint of the racial hierarchy. Because, this outcome occurs to children living in an often natal household, I can identify the race of both biological parents for many cases. This identification allows me to compare the likelihood of grade retention for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. ==PITHY SUMMARY==

@@

-   The biracial baby boom - why we should care about outcomes for this particular group.

-   What should we expect about the outcomes for biracials relative to their monoracial peer groups?

    -   In each case, our interest is in the outcomes for the biracial group relative to the two constituent groups that their parents identify with. So for example, how to the outcomes of a Black/White individual compare to monoracial Black and monoracial White individuals?
    -   The most direct pathway for understanding these outcomes is via how such individuals are classified by those in positions of power and control, such as teachers, administrators, employers, etc. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups, they will receive the same treatment by individuals and either reap similar benefits or suffer similar discrimination and penalties.
    -   Specifically, I consider two long-standing norms of *hypodescent* and *hyperdescent* in understanding possible placement of biracial individuals (cite F James Davis). Hypodescent indicates a norm in which individuals of mixed racial ancestry are classified as members of the lower status group within a racial hierarchy. In the US, hypodescent is canonically tied to the "one-drop rule", a norm in which those of part-Black ancestry were simply viewed as Black, irrespective of other ancestry (CITATIONS). More generally, if hypodescent is the common practice, we would expect biracial individual's outcomes to be closest those of the monoracial constituent group with the lower outcomes.
    -   Hyperdescent refers to the opposite situation in which a mixed race individual is classified as a member of the higher status group. Because hyperdescent blends individuals into the often unmarked dominant group, it is often more difficult to see. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming membership with another group (most notably Whites). In the case of outcomes, a pattern of hyperdescent would be indicated if biracial individual's outcomes are closest to those of the monoracial constituent group with the higher outcomes.
    -   Discuss work on the contemporary strength of these norms?
    -   The expectations of hypodescent/hyperdescent are that the biracial respondent will have outcomes close to one of the monoracial constituent groups. Another possibility is that biracial respondents will have outcomes intermediate between their monoracial constituent groups. There could be several reasons for this:
        -   The racial ambiguity of biracial respondents may protect them from the worst stigmatization and discrimination in social settings, but by the same coin it may allow them to benefit less from membership in advantaged groups. Thus, a greater recognition of their "blended" ancestry may place them in a halfway position.
        -   Even in cases where hypodescent and hyperdescent are still norms, physical differentation such as skin tone may provide advantages and disadvantages among members who are classified in the same way. We know that skin tone makes a difference even for monoracial Blacks, so the one-drop rule while providing a single classification does not create uniform experiences for all Blacks. (==CITE MONKS WORK HERE I THINK==).
        -   Successful outcomes within social systems depend on a variety of non-material, symbolic resources (e.g. social and cultural capital) that we rarely completely capture in our measurements. Some of the differences in outcomes across racial groups may reflect differences in those resources (CITATIONS?). Given that biracial children have parents belonging to two different races, they may possess such resources that are roughly half-way between the two constituent groups on average, and thus will place them roughly in between the groups. In effect, the in-between outcomes for biracial children may loosely reflect the "black box" package of differences across racial groups that a racial dummy variables stands in for in our research practices.
        -   Two other possibilities don't suggest the actual placement, but rather may modify it in the face of other prevalent patterns. Hybridity itself may be a boon or a burden on biracial individuals that affects their outcomes in, respectively, positive or negative ways.
            -   On the positive side of the ledger, hybridity may enabled a larger cultural "toolkit" that allows mixed race individuals to more easily navigate a variety of social situations. On the negative side of the ledger, the longstanding "marginal man" hypothesis suggests that the dissonance of living in two different worlds may lead to negative mental health outcomes for mixed race individuals, which could spill over into other types of outcomes.

-   In all of this, its crucially important to understand how other resources affect our understanding of the placement of biracial individuals. Biracial individuals are the products of interracial unions and individuals are not randomly selected into such unions. This can lead to underlying differences in the socioeconomic resources that biracial individuals possess relative to their monoracial peers. However, this selection can be quite complex. (DISCUSS CHENG AND LIVELY HERE?)

    -   Status exchange suggests that members of lower status racial groups will be positively selected into interracial unions, while higher status racial groups will be negatively selected. How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena, the gender balance in a given form of interracial unions, and how the gender of parents affects outcomes. Status exchange may also not be present in all forms of interracial unions.

-   Review previous studies on outcomes

    -   Campbell
    -   Herman
    -   Holloway/Ellis stuff
    -   Udry? - also related stuff by Cheng? How does Cheng and Powell fit in here?
    -   Do some more bibliographic work, particularly forward searching to see if I am missing something big here.
    -   Limitations of this work
        -   Small sample sizes lead to inconsistent estimates in most of the survey data.
        -   In many cases, impossible to understand outcomes for part-Latino biracials.

-   Why grade retention as an outcome?

    -   Grade retention is a relatively common practice. Discuss overall trends in grade retention and connection to educational reforms.
    -   Grade retention has negative effect on later outcomes, even controlling for cognitive differences (CITE WORK HERE).
    -   Clear inequalities by race in practices of grade retention as well as by SES (CITE WORK).
    -   In effect, grade retention may be for many individuals the first encounter with the wider stratification system that sorts individuals and helps to perpetuate inequality.

-   What I do and how it improves our understanding.

@@

# Conclusions

-   Big claim: The results are consistent with a new "regime of hyperdescent" operating across all racial groups. This contributes to a muddying of racial distinctiveness via hybridity, while at the same time isolating monoracial members of the most stigmatized racial groups, consistent with arguments like Bonilla-Silva's Latin Americanization.

# Acknowledgments {.appendix}

I am grateful to...

# References {.appendix}
